---
title: "SimuNet - First steps"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparing the field
## Data import

SimuNet's networks are simulated relying mostly on two aspects: 

*  a weighted adjacency matrix `Adj`:  
_filled with integers representing the number of times two nodes have been observed associated_  
*  and a sampling effort `total_scan`:  
_representing the number of times the edges has been sampled to fill in `Adj`_  
\

A user willing to run simulations with SimuNet can import an `Adj` matrix and `total_scan` sampling effort by:  

*  randomly generate an `Adj` matrix and `total_scan`:  
```{r random init}
set.seed(42)
library(SimuNet)

# for a random directed matrix
n <- 5;nodes <- letters[1:n];
total_scan <- round(runif(1,0,100))
total_scan
Adj <- matrix(data = 0L,nrow = n,
              ncol = n,
              dimnames = list(nodes,nodes)
)
Adj[non.diagonal(Adj)] <- round(runif(n*(n-1),0,total_scan))
Adj

```
*  import them from their real-life observations:  
```{r manuual init}
# create... 
Adj <- matrix(c(0,12,34,56,78,
                91,0,23,45,67,
                89,12,0,34,56,
                78,91,23,0,45,
                67,89,12,23,0),
              ncol = 5,byrow = TRUE,
              dimnames = list(1:5,1:5)
)
total_scan <- 100
Adj
# ... or import a matrix using your favorite method (e.g. read.csv)

```

*  find such data from the literature,  
*  or get them from the special case of the [Animal Social Network Repository](https://bansallab.github.io/asnr/)  

```{r asnr import as matrix}
Adj <- import_from_asnr("Mammalia",
                        "kangaroo_proximity_weighted",
                        output = "adjacency",type = "upper")
total_scan <- 241 # number of observations reported by the author
# data retrieved from (Grant, 1973)
# DOI: 10.1016/S0003-3472(73)80004-1
Adj
```
## Familiarizing with the data  
Our last example represented a social network of a group of kangaroo, the weights of which representing how many times each pair of kangaroos was observed in association, after 241 observations, or "scans".  
\  
The user can reconstruct an [igraph](https://igraph.org/r/) network (or graph) object from igraph's set of functions, but SimuNet's custom asnr importer can also output such an object directly by specifying the output (among "graph" and "adjacency"):  
```{r plotting original Adj}
G <- import_from_asnr("Mammalia",
                      "kangaroo_proximity_weighted",
                      output = "graph",type = "upper")
G
G.deg <- igraph::degree(G)
G.E <- igraph::E(G)$weight
plot(G,
     vertex.size = G.deg * 1.2,
     vertex.label = NA,
     edge.width = 15 * (G.E - min(G.E)) /
       (max(G.E) - min(G.E))
)
```
  
Before jumping into SimuNet, let's set one igraph layout to reuse later, for comparison purposes:  
```{r plotting set layout Adj}
G.layout <- igraph::layout_with_fr(G)
plot(G,
     vertex.size = G.deg * 1.2,
     vertex.label = NA,
     edge.width = 15 * (G.E - min(G.E)) / (max(G.E) - min(G.E)),
     layout = G.layout
)
```
  
# Introducing SimuNet  
## Basic functionalities  
SimuNet's goal is to randomly generate weighted (social) networks by simulating how each final weight has been in fact obtained after summing binary events.  
  
INSERT INFOGRAPHIC?
