---
title: "General approach"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## From theory to empiricism
### What we got so far
SimuNet generate network simulation in the form of `scanList`s.

By default, `scanList`s are of `scanList.type` `"theoretical"`: generated edges are drawn from the
posterior distribution obtained from [Bayesian inference](bayesian_inference.html), nothing more.
Therefore, edge weight distribution will follow an unaltered Beta-binomial distribution inferred
from inputted observed data. See more on the article on [network uncertainty](uncertainty.html)!

```{r setup}
library(SimuNet)
# TODO: insert a demo graph matching simulations with their expected beta-binomial distribution
```

### Experimental design
But `SimuNet` has been conceptualized to embed a way to _also_ simulate empirical phenomenon. These
can include:

* [**empirical sampling**](sampling.html): an observer relies on a method that can lead to data
diverging from their theoretical counterparts, due to random error, missing data, etc. SimuNet
includes way to simulate two main sampling methods used in animal social network analysis:
    * group-scan sampling, and
    * focal(-scan) sampling
* [**experimental manipulations**](experiments.html): these can represent abstract or artificial
modifications applied to the theoretical network in order to see the impact of such manipulations.
For instance, how is the resulting network impacted when the most peripheral node is removed (see
`remove_mostPeripheral()`)?

The general approach is based on the concept of experimental designs, that consists in performing
multiple "elementary" manipulations:

* Such manipulations to be done on a theoretical `scanList` can be combined, chained, and tweaked at
will.
* `SimuNet` provides "building blocks" of experimental design (see `design_exp()`), and will include
more in the future
* User can also write their own `scanList` manipulations to suits their specific needs!

```{r exp}
# TODO: insert a schematic
```

### Designing experiments
Once the user knows what sequence of manipulations defines the experimental design, the function
`design_exp()` can be used to generate a `expDesign` object:

* To do so, pass several functions in the desired order as arguments of `design_exp()`
* `design_exp()` returns an object of class `expDesign`, which main argument contain the sequence of
function to apply to a `scanList`. `design_exp()` internally relies on `purrr::compose()`
* You can also add manipulations to an existing `expDesign` object by passing it along other
functions in `design_exp()`. `design_exp()` accepts any combinations of functions and expDesign
objects!

### Running experiments
`expDesign` objects can be used as arguments in `simunet()` and `perform_exp()`

* When used in the former, the simulation outputs directly an empirical `scanList`, but the
intermediate theoretical `scanList` is stored in the attributes list `attrs` under the name
`theoretical.scanList`.
* The latter takes a theoretical `scanList` as an argument as well, and perform the inputted
`expDesign` to it. The returned `scanList` similarly stores the original in `attrs`

Users can also input several `expDesign` objects in one call of either `simunet()` and
`perform_exp()` to run several experimental design on a given theoretical `scanList`, allowing for
easier comparison:
* returned object is then a list of `scanList`s, aptly named `sLlist` objects.
* provided building blocks are written to seamlessly be applied to a `scanList` or `sLlist`, but
user-defined functions could benefit from using `lapply()` (e.g. `lapply(scanList,my_fun)`)

```{r expDesign}
# TODO: insert a schematic
```

### Thinking in experimental designs
Experimental design's philosophy can see inspiration from:

* the data manipulation philosophy used in tidy packages like
[`dplyr`](https://dplyr.tidyverse.org/) (manipulations and building blocks can even be piped!),
* as well as the concept of image and effect layers used in graphic design

We believe that with this approach based on experimental design, `SimuNet` can be used as a
framework to assess network uncertainty resulting from both the Bayesian framework adopted, as well
as under due to empirical matters.
