---
title: "First network simulations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let's first retrieve the kangaroo social network we [imported previously](importing_networks.html):
```{r setup}
library(SimuNet)

Adj <- import_from_asnr("Mammalia",
                        "kangaroo_proximity_weighted",
                        output = "adjacency",type = "upper")
samp.effort <- 241L # number of observations reported by the authors
# data retrieved from (Grant, 1973)
# DOI: 10.1016/S0003-3472(73)80004-1
Adj           # note that here, the suitable `mode` is driven by `type`
samp.effort
```

This represents an undirected network in the form of an upper-triangular matrix

## Running a simulation
Simulations are obtained via this package's main function `simunet()`. Let us see the detail of its
inputs:

```{r first_sim}
set.seed(42)

sL1 <- simunet(
  Adj = Adj,                 # the first 3 arguments represent what has been observed and will 
  samp.effort = samp.effort, # serve the purpose of determining a suitable distribution for
  mode = "upper",            # each edge presence probability.
        
  n.scans = 10               # this is how many scans should be _simulated_, which can be
)                            # different from the sampling effort samp.effort that was used
                             # to obtain Adj.
sL1
```
We obtain a 17 x 17 x 10 array of 10 binary adjacency matrices, the 10 scans.

## Running more simulations
Simulating more `scanList`s from the same `Adj` and `samp.effort` will lead not only to different
scans:
```{r second_sim}
sL2 <- simunet(Adj,samp.effort,"upper",n.scans = 10)
sL2

identical(sL1,sL2)
```

but also internally a different edge presence probability matrix. 

These matrices are stored in `scanList` their attribute list `attrs`, as the `edge.Prob` attribute,
and can be accessed via `attrs()` in a similar fashion than base `attr()`.
```{r edge.Probs}
attrs(sL1,"edge.Prob")
sL2 |> attrs("edge.Prob") # most of `SimuNet`'s functions have been written to allow for piping
```

## Collapsing the binary adjacency matrix sequence into a weighted adjacency matrix
### Summing `scanList`
`scanList`'s 3D arrays can be summed over their 3rd dimension via `sum_scans()`:
```{r sum_firsts}
sum_scans(sL1)
sL2 |> sum_scans()
```

For future comparison purpose, let us define a function calculating the correlation between the
edges of two triangular matrices:
```{r upper_cor}
upper_cor <- function(X,Y) {
  cor(X[upper.tri(X)],Y[upper.tri(Y)])
}
```

Now, we run a new simulation containing as many scans as the input sampling effort to see how the
simulation compares to `Adj`
```{r sum_comparison}
sL3.Adj <- simunet(Adj,samp.effort,"upper",n.scans = samp.effort) |> sum_scans()

sL3.Adj |> attrs("Adj")
sL3.Adj

upper_cor(Adj,sL3.Adj)

plot(Adj[upper.tri(Adj)],sL3.Adj[upper.tri(sL3.Adj)],
    xlab = "Original network edges' weights (Adj)",
    ylab = "Simulated network edges's weights"
)
abline(0,1,col = "red",lty = "dashed")
```

### Scaling `scanList`
To compare `scanList` of different size, another "collapsing" function, `scale_scans()`, can be
applied to `scanList` or their collapsed sums. `scale_scans()` divide the sum of binary edges by the
number of time they were sampled (see more about sampling with `customize_sampling()` and [empirical
designs](experiments.html)).

In the case of theoretical scanList, this means dividing the obtained weighted adjacency matrix by
`n.scans`.

```{r scale_firsts}
scale_scans(sL1)
sL2 |> scale_scans()
scale_scans(sL3.Adj)
```

One can look at how the correlation with a scaled `Adj` evolve with increasing `n.scans`
```{r scale_details}
scaled.list <- 
  lapply(
    round(seq(10,250,length.out = 200)),
    function(n) simunet(Adj,samp.effort,"upper",n) |> scale_scans()
  )

scaled.list |>
  sapply(upper_cor,Y = Adj / samp.effort) |>
  plot(xlab = "n.scans",ylab = "edge correlation")
```
