---
title: "Importing Network data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Before starting with network simulations, `SimuNet` needs networks (e.g. from real-life) to
"inspire" its internal simulation probabilities. The approach is described in the article on the
[underlying Bayesian framework](bayesian_framework.html), and is described "code-wise" in the
`draw_edgeProb()`'s doc.

## Data import
`SimuNet`requires two types of data: 

* a weighted adjacency matrix `Adj`: matrix of the number of times (integers) two nodes have been
observed associating or interacting
* a sampling effort `samp.effort`: number of times the edges haVe been sampled to observe `Adj`

A user willing to run simulations with SimuNet will import an `Adj` matrix and `samp.effort`.

Care regarding the `mode` of `Adj`  - in the `igraph` sense (see `igraph::graph.adjacency()`) - is
also required. In short, `igraph` defines the `mode`s:

* `"directed"` for directed networks (asymmetrical matrices) where $Adj[i,j] = 1$ does not have the
same meaning as $Adj[j,i] = 1$
* `"undirected"` for undirected networks (symmetrical matrices) where $Adj[i,j] = 1$ has the same
meaning as $Adj[j,i] = 1$.    
`igraph` actually includes multiple approaches to build undirected networks when an asymmetrical
matrix is inputted:

    * `"max"`, which takes the $max(Adj[i,j],Adj[j,i])$ for both edges. This is actually the `mode`
    used when `mode = "undirected"`
    * `"min"`, which takes the $max(Adj[i,j],Adj[j,i])$ for both edges
    * `"plus"`, which takes $Adj[i,j] + Adj[j,i]$ for both edges
    * `"upper"`, for upper-triangular matrices
    * `"lower"`, for lower-triangular matrices

For unambiguous undirected networks, we recommend using triangular matrices (`mode = "upper"` or
`mode = "upper"`).

### Creating or importing `Adj` and `samp.effort`
Any methods leading to such a square integer matrix - triangular or not - is suitable as `Adj`. For
`samp.effort`, simply assign an integer to a variable.

#### randomly generate an `Adj` matrix and `samp.effort`:  
```{r random init}
library(SimuNet)
set.seed(42)

# for a random directed matrix
n <- 5L
nodes <- letters[1:n]
samp.effort <- runif(1,0,100) |> round()
Adj <- matrix(data = 0L,nrow = n,
              ncol = n,
              dimnames = list(nodes,nodes)
)
Adj[non.diagonal(Adj)] <- runif(n * (n - 1),0,samp.effort) |> round()

Adj          # note that the suitable `mode` here would be "directed"
samp.effort
```

#### import them manually
from real-life observations:  
```{r manual init}
# input data manually... 
Adj <- matrix(c( 0,12,34,56,78,
                 0, 0,23,45,67,
                 0, 0, 0,34,56,
                 0, 0, 0, 0,45,
                 0, 0, 0, 0, 0),
              ncol = 5,byrow = TRUE,
              dimnames = list(1:5,1:5)
)
samp.effort <- 100L

Adj           # note that the suitable `mode` here would be "upper"
samp.effort
# ... or import a matrix using your favorite method (e.g. read.csv)
```

Or with such data from the literature

#### Import them from the [Animal Social Network Repository](https://bansallab.github.io/asnr/)  
A helper function, `import_from_asnr()`, can be used to interact with the
[ASNR](https://bansallab.github.io/asnr/)
```{r asnr import as matrix}
Adj <- import_from_asnr("Mammalia",
                        "kangaroo_proximity_weighted",
                        output = "adjacency",type = "lower")
samp.effort <- 241L # number of observations reported by the author
# data retrieved from (Grant, 1973)
# DOI: 10.1016/S0003-3472(73)80004-1
Adj           # note that here, the suitable `mode` is driven by `type`
samp.effort
```

That's it! With `Adj`, `samp.effort` and knowing the suitable `mode`, you can start using
`simunet()` to [generate social network simulations](first_simulations.html)!

Hereafter, we only showcase general network considerations about converting, importing and plotting
networks.

## Familiarizing with the data  
Our last example represented a social network of a group of kangaroo, the weights of which
representing how many times each pair of kangaroos was observed in association, after 241
observations, or `"scans"`.

The user can reconstruct an [igraph](https://igraph.org/r/) network (or graph) object from
`igraph::graph.adjacency()`, but `SimuNet`'s `import_from_asnr()` can also output igraph object
directly by specifying the `output` (among `"graph"` and `"adjacency"`, with support for partial
matching)

From there, it is easy to plot a visualization of the network:
```{r plotting original Adj}
# from Adj...
G <- igraph::graph.adjacency(Adj,mode = "lower",weighted = TRUE)
# ... or directly imported
G <- import_from_asnr("Mammalia",
                      "kangaroo_proximity_weighted",
                      output = "graph",type = "lower")

G.deg <- igraph::degree(G)
G.E <- igraph::E(G)$weight
plot(G,
     vertex.size = G.deg * 1.2,
     vertex.label = NA,
     edge.width = 15 * (G.E - min(G.E)) /
       (max(G.E) - min(G.E))
)
```

The conversion can be achieve the other-way around:
```{r network to Adj}
G.Adj <- igraph::get.adjacency(G,type = "lower",attr = "weight")
G.Adj

identical(G.Adj |> as.matrix(),Adj)
```
